09-understanding_lime

### 2019-08-20 README.Rmd
updates since last meeting

@@

- work on first draft of paper
- read through papers giving an overview of the current explanation methods

@@

- LIME diganostics  
    - turn repetitive code into functions
    - try out different weights
    - change distance matrix in seriation to be 5-# of variables shared between two rows
    - try using distance matrix computed on the feature value with seriation
    - add in feature selection methods to LIME input options
    - think of a way to compute consistency across top two features
    - Siggi suggests refitting the RF model to the perturbations and then continuing with LIME with the RF predictions from the new model - this may help to understand if the problems are due to the sampling procedure or LIME itself
    - he also suggested looking into SMOTE for dealing the inbalance in the classes with sampling
    - compare the simple models based on different number of bins using an F-test
    - look into computing a diversity or consistency measure for the sensitivity analysis
    - include a penalty for the number of parameters when choosing bins
    - look at the AUC after binning
    - compute a likelihood ratio prob TRUE / prob FALSE from the LIME ridge regression
    - try visualizing the features from the test data using dimension reduction and
  coloring them by variables suggested to be important by lime
- explainer models
    - look into iml (https://www.youtube.com/watch?v=jP6Rg13PEkQ)
    - look more into Molnar and his adviser Bernd
    - look into [MAPLE](https://blog.ml.cmu.edu/2019/07/13/towards-interpretable-tree-ensembles/)
- understanding LIME
    - read through LIME papers again
    - finish journal on understanding lime (look at the RF in the local region lime is considering to see if it agrees with the lime explanations)
    - try fitting LASSO logistic model and leave one out approach (for multicollinarity)
    - assess lime results on logistic regression models
    - think about how the argmin is accomplished by the LIME R package
    - try retriculate to apply python version of lime
    - look into literature on binning methods
    - think about why R^2 would be better for some binning methods
    - read new paper on Anchor
- simulation with RF
    - think about how to implement a simulation with a random forest
    - look back at 601 notes from Kaiser for model assessment
    - look at Hadley's 'removing the blindfold' for ideas
    - read book Dr. Dixon lent me on sensativity analyses
- possible improvements to LIME
    - determine the best number of bins to use for each variable
    - try out subsampling idea
- random forests confidence intervals
    - look into papers by Giles Hooker and Lucas Mentch
    - Dr. Nettleton paper

@@

Descriptions of the materials in this respository are listed below.

@@

- `00-objectives_and_ideas`: background of the research project, goals of the project, concerns with LIME, and ideas to try
- `01-training_and_testing_data`: information on the Hamby data, understanding the raw data, and cleaning of the training and testing data
- `02-plotting_training_and_testing_data`: exploratory plots of the training and testing data
- `03-computation_issues`: documentation of some of the computation issues I ran into during the project
- `04-lime_algorithm_and_proposed_methods`: rough draft of the lime algorithm as implemented in the lime R package and proposed methods for improving the lime algorithm
- `05-applying_lime_to_rtrees`: application of lime and sensativity analysis for many different input options
- `06-assessing_lime_explanations`: assessment of the lime explanations
- `07-logistic_regression`: logistic regressions fit to the bullet data and application of lime to the logistic regression
- `08-iris_comparison`: applying LIME to a random forest fit to the iris data
- `09-understanding_lime`: attempts to visualize the results from LIME to understand how LIME works
- `10-other_explainers`: trying out other model explainers

>>>>

- Paper
    - I found ridge regression models in Python code (see journal 04)
    - Show the plots using the ridge regression in journal 09
    - What should I do about computing the generalized R2 value?
    - Show this paper: https://arxiv.org/pdf/1806.07498.pdf
    - Can you help to show me your vision with the minipage format?
    - I could not find a locality metric output from Python. There was a section in their vignettes that seemed to be referring to some sort of assessment about the models, but I'm not sure what was happening. https://marcotcr.github.io/lime/tutorials/Tutorial%20-%20continuous%20and%20categorical%20features.html
- Graphics Group
    - email list (should I be added as an admin?)
    - send out another email and plan for first week?
    - updates on schedule since last meeting

@@

- keep the LIME procedure at a higher level in the paper
- change to computing the $\mathcal{L}$ metric from the paper
- add figure descriptions in both caption and text
- go back through LIME procedure and update with new understanding (such as d' can be larger than d)
- add an assert_that statement in the code to check if the LIME package is changing
- read through papers giving an overview of the current explanation methods

@@

- LIME diagnostics  
    - turn repetitive code into functions
    - try out different weights
    - change distance matrix in seriation to be 5-# of variables shared between two rows
    - try using distance matrix computed on the feature value with seriation
    - add in feature selection methods to LIME input options
    - think of a way to compute consistency across top two features
    - Siggi suggests refitting the RF model to the perturbations and then continuing with LIME with the RF predictions from the new model - this may help to understand if the problems are due to the sampling procedure or LIME itself
    - he also suggested looking into SMOTE for dealing the imbalance in the classes with sampling
    - compare the simple models based on different number of bins using an F-test
    - look into computing a diversity or consistency measure for the sensitivity analysis
    - include a penalty for the number of parameters when choosing bins
    - look at the AUC after binning
    - compute a likelihood ratio prob TRUE / prob FALSE from the LIME ridge regression
    - try visualizing the features from the test data using dimension reduction and
  coloring them by variables suggested to be important by lime
- explainer models
    - look into iml (https://www.youtube.com/watch?v=jP6Rg13PEkQ)
    - look more into Molnar and his adviser Bernd
    - look into [MAPLE](https://blog.ml.cmu.edu/2019/07/13/towards-interpretable-tree-ensembles/)
- understanding LIME
    - read through LIME papers again
    - finish journal on understanding lime (look at the RF in the local region lime is considering to see if it agrees with the lime explanations)
    - try fitting LASSO logistic model and leave one out approach (for multicollinarity)
    - assess lime results on logistic regression models
    - think about how the argmin is accomplished by the LIME R package
    - try retriculate to apply python version of lime
    - look into literature on binning methods
    - think about why R^2 would be better for some binning methods
    - read new paper on Anchor
- simulation with RF
    - think about how to implement a simulation with a random forest
    - look back at 601 notes from Kaiser for model assessment
    - look at Hadley's 'removing the blindfold' for ideas
    - read book Dr. Dixon lent me on sensitivity analyses
- possible improvements to LIME
    - determine the best number of bins to use for each variable
    - try out subsampling idea
- random forests confidence intervals
    - look into papers by Giles Hooker and Lucas Mentch
    - Dr. Nettleton paper

@@

Descriptions of the materials in this repository are listed below.

@@

- `00-objectives_and_ideas`: background of the research project, goals of the project, concerns with LIME, and ideas to try
- `01-training_and_testing_data`: information on the Hamby data, understanding the raw data, and cleaning of the training and testing data
- `02-plotting_training_and_testing_data`: exploratory plots of the training and testing data
- `03-computation_issues`: documentation of some of the computation issues I ran into during the project
- `04-lime_algorithm_and_proposed_methods`: rough draft of the lime algorithm as implemented in the lime R package and proposed methods for improving the lime algorithm
- `05-applying_lime_to_rtrees`: application of lime and sensitivity analysis for many different input options
- `06-assessing_lime_explanations`: assessment of the lime explanations
- `07-logistic_regression`: logistic regressions fit to the bullet data and application of lime to the logistic regression
- `08-iris_comparison`: applying LIME to a random forest fit to the iris data
- `09-understanding_lime`: attempts to visualize the results from LIME to understand how LIME works
- `10-other_explainers`: trying out other model explainers

### 2019-08-20 journals/04-lime_algorithm_and_proposed_methods/04-lime_algorithm_and_proposed_methods.Rmd
updates since last meeting

@@

```{r}
# Load in the training data (Hamby Data 173 and 252)
hamby173and252_train <- read.csv("../../data/hamby173and252_train.csv")
```

>>>>

# Python Package

I have not spent a lot of time going through the Python package code, but I wanted to check to see if it uses a ridge regression for the feature selection and the explainer model. This [article](https://towardsdatascience.com/how-to-perform-lasso-and-ridge-regression-in-python-3b3b75541ad8) describes that ridge regression can be fit in Python using the sklearn "package". These images below of the code from the Python lime [GitHub repository](https://github.com/marcotcr/lime/blob/master/lime/lime_base.py) seem to imply that it is using ridge regression for both of these steps.

![](../../figures/python-ridge1.png)

![](../../figures/python-ridge2.png)

![](../../figures/python-ridge3.png)

@@

```{r}
# Load in the training data (Hamby Data 173 and 252)
hamby173and252_train <- read.csv("../../../data/hamby173and252_train.csv")
```

### 2019-08-20 journals/09-understanding_lime/09-understanding_lime.Rmd
updates since last meeting

```{r}
library(furrr)
library(future)
library(lime)
library(randomForest)
library(tidyverse)

@@

# The Data

@@

```{r}
# Load in the training data (Hamby Data 173 and 252)
hamby173and252_train <- read.csv("../../data/hamby173and252_train.csv")

# Load in the testing data (Hamby Data 224 Sets 1 and 11)
hamby224_test <- read.csv("../../data/hamby224_test.csv")
```

@@

# Random Forest Model

@@

# Applying LIME

@@

```{r}
# Create or load the lime and explain objects for the main effects logisitic regression model
if(!file.exists("./data/lime_explain_sub.rds")) {

@@

```{r}
create_bin_df <- function(bin_cut) {
 data.frame(quantile = c("0%", "25%", "50%", "75%", "100%"), 
            bin_cut) 
}

@@

# Visualizations of Data, Model, and LIME

```{r}
(case1 <- lime_explain_sub$explain %>% filter(case == 1))
```

```{r}
ggplot() +
  geom_point(data = hamby173and252_train %>% filter(samesource == FALSE),
             aes(x = rough_cor, y = ccf),
             color = "darkgrey",
             alpha = 0.2) +
  geom_point(data = hamby173and252_train %>% filter(samesource == TRUE),
             aes(x = rough_cor, y = ccf),
             color = "darkorange") + 
  geom_vline(xintercept = bin_cuts$rough_cor) + 
  geom_hline(yintercept = bin_cuts$ccf) +
  theme_minimal()
```

```{r}
ggplot() + 
  geom_point(data = hamby224_test_subset, 
             aes(x = rough_cor, y = ccf, color = rfscore),
             alpha = 0.5) + 
  scale_color_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  geom_point(data = lime_explain_sub$explain %>% 
               filter(case == 1) %>%
               select(feature, feature_value) %>%
               spread(key = feature, value = feature_value),
             aes(x = rough_cor, y = ccf)) +
  geom_vline(xintercept = bin_cuts$rough_cor) + 
  geom_hline(yintercept = bin_cuts$ccf) + 
  theme_minimal()
```

>>>>

This journal is a place where I create visuals to try to understand how lime works.

```{r}
library(cowplot)
library(furrr)
library(future)
library(glmnet)
library(gower)
library(lime)
library(randomForest)
library(tidyverse)

@@

# Generated Data

The images in this section were created using simulated data that I knew had a local trend around the point of interest. I consider both a linear regression and a ridge regression as the explainer model.

### Data and Model

This code generates the data and predictions from a hypothetical complex model.

```{r}
# Generate features and predictions from a hypothetical complex model
set.seed(20190624)
lime_data <- data.frame(feature1 = sort(runif(250, 0, 1)),
                        feature2 = sample(x = 1:250, 
                                          size = 250, 
                                          replace = FALSE)) %>%
  mutate(prediction = if_else(feature1 >= 0 & feature1 < 0.1, 
                              (0.3 * feature1) + rnorm(n(), 0, 0.01), 
                      if_else(feature1 >= 0.1 & feature1 < 0.3, 
                              rbeta(n(), 1, 0.5), 
                      if_else(feature1 >= 0.3 & feature1 < 0.5, 
                              sin(pi* feature1) + rnorm(n(), 0, 0.5),
                      if_else(feature1 >= 0.5 & feature1 < 0.8, 
                              -(sin(pi* feature1) + rnorm(n(), 0, 0.1)) + 1,
                      if_else(feature1 >= 0.8 & feature1 < 0.9, 
                              0.5 + runif(n(), -0.5, 0.5), 
                              0.5 + rnorm(n(), 0, 0.3)))))))
```

```{r}
# Specify a prediction of interest
prediction_of_interest <- data.frame(feature1 = 0.07, 
                                     feature2 = 200,
                                     prediction = 0.05, 
                                     color = factor("Prediction of Interest"))
```

```{r}
# Compute the distance between the prediction of interest and 
# all other observations
lime_data$distance <- (1 - gower_dist(x = prediction_of_interest, 
                                      y = lime_data))^20
```

### Linear Regression Explainer

These visualizations are based on a weighted linear regression model used as the explainer.

```{r}
# Fit the interpretable explainer model (weighted linear regression)
linear_explainer <- lm(formula = prediction ~ feature1 + feature2, 
                       data = lime_data, 
                       weights = distance)
```

```{r}
# Extract the coefficients from the explainer
linear_b0 <- coef(linear_explainer)[[1]]
linear_b1 <- coef(linear_explainer)[[2]]
linear_b2 <- coef(linear_explainer)[[3]]
```

```{r}
# Predictions versus feature 1
linear_plot_feature1 <- ggplot(data = lime_data,
                        mapping = aes(x = feature1, 
                                      y = prediction, 
                                      size = distance)) + 
  geom_point(color = "grey30",
             alpha = 0.75) + 
  geom_abline(intercept = linear_b0 +
                linear_b2*prediction_of_interest$feature2, 
              slope = linear_b1, 
              size = 1) +
  geom_point(data = prediction_of_interest,
             mapping = aes(x = feature1, y = prediction),
             fill = "#EFB084",
             color = "black",
             size = 5, 
             shape = 23) +
  theme_linedraw(base_family = "Times") +
  labs(x = "Feature 1", 
       y = "Black-Box Prediction") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")
```

```{r}
# Predictions versus feature 1
linear_plot_feature2 <- ggplot(data = lime_data,
                        mapping = aes(x = feature2, 
                                      y = prediction, 
                                      size = distance)) + 
  geom_point(color = "grey30",
             alpha = 0.75) + 
  geom_abline(intercept = linear_b0 + 
                linear_b1*prediction_of_interest$feature1,
              slope = linear_b2,
              size = 1) +
  geom_point(data = prediction_of_interest,
             mapping = aes(x = feature2, y = prediction),
             fill = "#EFB084",
             color = "black",
             size = 5, 
             shape = 23) +
  theme_linedraw(base_family = "Times") +
  labs(x = "Feature 2", 
       y = "Black-Box Prediction") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")
```

```{r}
# Plot that creates a size legend
size_legend_plot <- lime_data %>%
  gather(key = feature, value = value, 1:2) %>%
  ggplot(mapping = aes(x = value, 
                       y = prediction,
                       size = distance)) + 
  geom_point(color = "grey30",
             alpha = 0.75) +
  labs(size = "Weight") +
  theme_linedraw(base_family = "Times") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

# Extract the legend
size_legend <- get_legend(size_legend_plot)
```

```{r}
# Join the two feature plots
lime_linear_plots <- plot_grid(linear_plot_feature1, 
                               linear_plot_feature2, 
                               size_legend,
                               ncol = 3, 
                               rel_widths = c(0.4, 0.4, 0.1))
```

```{r}
# Create a plot to extract a legend for the prediction of interest
lime_legend_plot <- ggplot(data = prediction_of_interest, 
                           mapping = aes(x = feature1, 
                                         y = prediction, 
                                         fill = color)) +
  geom_point(size = 5, shape = 23) +
  scale_fill_manual(values = c("#EFB084")) +
  theme_classic(base_family = "Times") +
  labs(fill = "") +
  theme(legend.position = "bottom")

# Extract the legend
lime_legend <- get_legend(lime_legend_plot)
```

```{r}
# Join the title, plots, legend, and caption into one figure
lime_linear_concept <- plot_grid(lime_linear_plots, lime_legend,
                         ncol = 1,
                         rel_heights = c(0.9, 0.1))
```

This example shows that the explainer model is doing a good job of capturing the local trend.

```{r}
# View the plot
lime_linear_concept
```

### Ridge Regression Explainer

```{r}
# Fits the explainer model (weighted ridge regression)
ridge_explainer <- glmnet(x = lime_data %>% 
                            select(feature1, feature2) %>%
                            as.matrix(), 
                          y = lime_data %>% pull(prediction),
                          weights = lime_data$distance,
                          family = "gaussian", 
                          alpha = 0, 
                          lambda = 1, 
                          standardize = FALSE)
```

```{r}
# Extract the coefficients from the explainer
ridge_b0 <- ridge_explainer$a0[[1]]
ridge_b1 <- ridge_explainer$beta[1,1]
ridge_b2 <- ridge_explainer$beta[2,1]
```

```{r}
# Predictions versus feature 1
ridge_plot_feature1 <- ggplot(data = lime_data,
                        mapping = aes(x = feature1, 
                                      y = prediction, 
                                      size = distance)) + 
  geom_point(color = "grey30",
             alpha = 0.75) + 
  geom_abline(intercept = ridge_b0 +
                ridge_b2*prediction_of_interest$feature2, 
              slope = ridge_b1, 
              size = 1) +
  geom_point(data = prediction_of_interest,
             mapping = aes(x = feature1, y = prediction),
             fill = "#EFB084",
             color = "black",
             size = 5, 
             shape = 23) +
  theme_linedraw(base_family = "Times") +
  labs(x = "Feature 1", 
       y = "Black-Box Prediction") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")
```

```{r}
# Predictions versus feature 2
ridge_plot_feature2 <- ggplot(data = lime_data,
                        mapping = aes(x = feature2, 
                                      y = prediction, 
                                      size = distance)) + 
  geom_point(color = "grey30",
             alpha = 0.75) + 
  geom_abline(intercept = ridge_b0 + 
                ridge_b1*prediction_of_interest$feature1,
              slope = ridge_b2,
              size = 1) +
  geom_point(data = prediction_of_interest,
             mapping = aes(x = feature2, y = prediction),
             fill = "#EFB084",
             color = "black",
             size = 5, 
             shape = 23) +
  theme_linedraw(base_family = "Times") +
  labs(x = "Feature 2", 
       y = "Black-Box Prediction") + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")
```

```{r}
# Join the two feature plots
lime_ridge_plots <- plot_grid(ridge_plot_feature1, 
                              ridge_plot_feature2, 
                              size_legend,
                              ncol = 3,
                              rel_widths = c(0.4, 0.4, 0.1))
```

```{r}
# Join the title, plots, legend, and caption into one figure
lime_ridge_concept <- plot_grid(lime_ridge_plots, 
                                lime_legend,
                                ncol = 1,
                                rel_heights = c(0.9, 0.1))
```

This example shows that the explainer model is **not** doing a good job of capturing the local trend.

```{r}
# Preview the plot
lime_ridge_concept
```

# Bullet Data

### The Data

@@

```{r}
# Load in the training data (Hamby Data 173 and 252)
hamby173and252_train <- read.csv("../../../data/hamby173and252_train.csv")

# Load in the testing data (Hamby Data 224 Sets 1 and 11)
hamby224_test <- read.csv("../../../data/hamby224_test.csv")
```

@@

### Random Forest Model

@@

### Applying LIME

@@

```{r eval = FALSE}
# Create or load the lime and explain objects for the main effects logisitic regression model
if(!file.exists("./data/lime_explain_sub.rds")) {

@@

```{r eval = FALSE}
create_bin_df <- function(bin_cut) {
 data.frame(quantile = c("0%", "25%", "50%", "75%", "100%"), 
            bin_cut) 
}

@@

### Visualizations of Data, Model, and LIME

```{r eval = FALSE}
(case1 <- lime_explain_sub$explain %>% filter(case == 1))
```

```{r eval = FALSE}
ggplot() +
  geom_point(data = hamby173and252_train %>% filter(samesource == FALSE),
             aes(x = rough_cor, y = ccf),
             color = "darkgrey",
             alpha = 0.2) +
  geom_point(data = hamby173and252_train %>% filter(samesource == TRUE),
             aes(x = rough_cor, y = ccf),
             color = "darkorange") + 
  geom_vline(xintercept = bin_cuts$rough_cor) + 
  geom_hline(yintercept = bin_cuts$ccf) +
  theme_minimal()
```

```{r eval = FALSE}
ggplot() + 
  geom_point(data = hamby224_test_subset, 
             aes(x = rough_cor, y = ccf, color = rfscore),
             alpha = 0.5) + 
  scale_color_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  geom_point(data = lime_explain_sub$explain %>% 
               filter(case == 1) %>%
               select(feature, feature_value) %>%
               spread(key = feature, value = feature_value),
             aes(x = rough_cor, y = ccf)) +
  geom_vline(xintercept = bin_cuts$rough_cor) + 
  geom_hline(yintercept = bin_cuts$ccf) + 
  theme_minimal()
```
